---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sizeR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `sizeR` is to provide a collection of computational tools to analyze gene sizes within expression data or gene sets.

## Requirements

The `sizeR` package requires several packages from `tidyverse`. Either `tidyverse` can be installed and loaded, or these individual packages can be installed and loaded:

* `dplyr`
* `tidyr`
* `readr`
* `ggplot2`

## Installation

You can install the development version of `sizeR` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mjmccoy/sizeR")
```

## Example usage

This is series of examples which shows how to perform gene size enrichment analysis with either quantitative or categorical variables:

### Load libraries
```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(sizeR)
```

### Load user-specified gene lengths
```{r gene lengths, message=FALSE, warning=FALSE}
# Load user_specified gene lengths
lengths.df <- gene_lengths(filepath = "inst/extdata/hsap_GRCh38.p14_ensembl_release_110_gene_lengths.txt")
```

### Estimate gene size enrichment for quantitative variables
```{r quantitative variables, message=FALSE, warning=FALSE}
################################################
# Example 1: Gene size v. quantitative feature #
################################################

# Load data
data.df <- read_csv(file = "inst/extdata/example_expression_data.csv")

# Append gene lengths
data.df <- add_lengths(data.df, lengths.df)

# Binomial test
binomial.df <- binomial_test(data.df, feature_name = "log2FC")

# Binomial tile plot
binomial_tile_plot(binomial.df)
```

```{r fig.height=3, fig.width=14}
# Binomial bar plot
binomial_bar_plot(binomial.df)
```

### Estimate gene size enrichment for quantitative variables by sample
```{r quantitative variable by sample, fig.height=3, fig.width=12, message=FALSE, warning=FALSE}
##########################################################
# Example 2: Gene size v. quantitative feature by sample #
##########################################################

# Load data
data.df <- read_csv(file = "inst/extdata/example_expression_by_sample_data.csv")

# Append gene lengths
data.df <- add_lengths(data.df, lengths.df)

# Binomial test
binomial.df <- binomial_test(data.df, by_sample = T)

# Binomial tile plot
binomial_tile_plot(binomial.df, by_sample = T)
```

```{r fig.height=12, fig.width=12}
# Binomial bar plot
binomial_bar_plot(binomial.df, by_sample = T)
```

### Estimate gene size enrichment for categorical variables
```{r categorical variable, fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
##################################################
# Example 3: Gene size v. gene set (categorical) #
##################################################

# Load gene set data
data.df <- read_csv(file = "inst/extdata/example_gene_set_data.csv")

# Append gene lengths
data.df <- add_lengths(data.df, lengths.df)

# Binomial test
binomial.df <- binomial_test(data.df, categorical = T)

# Binomial tile plot
binomial_tile_plot(binomial.df, categorical = T)

# Binomial bar plot
binomial_bar_plot(binomial.df, categorical = T)
```
